Term Project: Q-Learning

### 2017-06-10 paper.tex
Updates to paper; small modifications to program to support testing.

\documentclass[12pt,a4paper]{article}
\usepackage[margin=1in]{geometry}
\usepackage{mathtools, qtree}
\usepackage{amssymb,amsmath, fancyhdr}
\usepackage{tikz}
\usetikzlibrary{automata,positioning}
\usepackage{listings}
\pagestyle{fancy}
\title{Term Project: Q-Learning}

@@

		\subsection{Instrumentation}

@@

	\section{Testing \& Results}

@@

			\subsubsection{Performance}

		\subsection{Data Collected}

		\subsection{Interpretation of Data}

		\subsection{Analysis of Algorithm}

@@

@@

	\pagebreak
	\begin{thebibliography}{9}
		\bibitem{RL}
		Melanie  Mitchell.
		\textit{Reinforcement Learning}.
		http://web.cecs.pdx.edu/~mm/MachineLearningWinter2017/RL.pdf
		\bibitem{alphago}
		Silver, D., A. Huang, Aja, C. Maddison, A. Guez, L. Sifre, G. van den Driessche, J. Schrittwieser, I. Antonoglou, V. Panneershelvam, M. Lanctot, S. Dieleman, D. Grewe, J. Nham, N. Kalchbrenner, I. Sutskever, T. Lillicrap, M. Leach, K. Kavukcuoglu, T. Graepel, D. Hassabis. 2016. "Mastering the Game of Go with Deep Neural Networks and Tree Search". Nature 529 (28 Jan): 484--503.
		\bibitem{kleisli}
		Monad laws.
		https://wiki.haskell.org/Monad\_laws
	\end{thebibliography}

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage[margin=1in]{geometry}
\usepackage{mathtools, qtree}
\usepackage{amssymb,amsmath, fancyhdr}
\usepackage{tikz}
\usetikzlibrary{automata,positioning}
\usepackage{listings}
\usepackage{graphicx}
\pagestyle{fancy}
\title{Term Project: Q-Learning}

@@

		\subsection{Instrumentation}\label{inst}

@@

	\section{Testing \& Results}\label{testingresults}

	\par In this section I'll discuss my process of testing my program, comparing it to a Python version, and collecting and interpreting the data it generated.

@@

			\par Testing for this program was done through manual inspection of outputs. When testing, I checked output values against expected values, and also used output values in assessment of ``expected behavior" of the algorithm in the context of my previously written working program.
			\par Of utmost importance was board generation, and agreement between the formal board parameters and the visual representation of the board. We can start small and manually check these values: \\
			\begin{lstlisting}[language=Haskell,numbers=none]
*Qlearn> b <- randBoard (4,4) 0.5
*Qlearn> b
Board ((4,4),[Can (2,1),Can (3,2),Can (1,2),Can (2,0),Can (3,1),Can (3,3),
Can (1,3)],Rob (0,0))
*Qlearn> showBoard b
[ ][.][.][.]
[.][.][ ][ ]
[ ][ ][.][.]
[o][ ][ ][ ]
			\end{lstlisting}\ \\

			\par Here we expect that the all the Can and Rob coordinates lie inside the board dimensions, i.e. $0 \leq i < dim$. We also expect there to be a number of cans approximately equal to the floating point value passed as the second argument to \texttt{randBoard}, in this case $0.5$. I use the word ``approximately" here because of two reasons: 1. integer values means an exact percentage may not be possible, and 2. the cans are randomly generated and duplicates may occur (in which case they are pruned out by the algorithm).
			\par Furthermore, we expect the display generated by \texttt{showBoard} to agree with the formal contents of the Board. Hand-verifying in small cases like these is relatively easy, but also slow and based on my own perceptive abilities. Given more time, automated testing could have been implemented to attain wider coverage.
			\par Keep in mind that the first value in a tuple represents its $i$ value, or location on the $y$-axis, and the second value represents its $j$ value, or location on the $x$-axis. This is slightly clunky from an array manipulation, computer science perspective, but fits nicely into the visual implementation of a grid as a list of lists. It also agrees with mathematical treatments of matrices.
			\par We can scale this up; the manual examination becomes more tedious but at least we have some assurance that the program is behaving as expected:\\

			\begin{lstlisting}[language=Haskell,numbers=none]
*Qlearn> b <- randBoard (8,8) 0.2
*Qlearn> b
Board ((8,8),[Can (1,4),Can (7,0),Can (2,0),Can (0,0),Can (1,5),Can (0,2),
Can (4,6),Can (6,2),Can (7,2),Can (4,0),Can (2,2),Can (0,3)],Rob (0,6))
*Qlearn> showBoard b
[.][ ][.][ ][ ][ ][ ][ ]
[ ][ ][.][ ][ ][ ][ ][ ]
[ ][ ][ ][ ][ ][ ][ ][ ]
[.][ ][ ][ ][ ][ ][.][ ]
[ ][ ][ ][ ][ ][ ][ ][ ]
[.][ ][.][ ][ ][ ][ ][ ]
[ ][ ][ ][ ][.][.][ ][ ]
[.][ ][.][.][ ][ ][o][ ]
			\end{lstlisting}\ \\

			\par Here we have 12 cans ($12.8$ being the exact figure for $(8 \times 8) \times 0.2$), and we can see that all coordinates fall within the given dimensions.
			\par We are going to ignore some edge cases here, such as board dimensions $(0,0)$, because such a case would be outside the problem domain -- i.e. it would make no sense and be of no use to train a Q-Learning agent on a board with no squares. 
			\par Additionally, I was able to use the \texttt{showBoard} function to verify that the agent's movements on the board were happening as expected. I won't reproduce the output here for space considerations, but one can see in the code appendix where the print statements facilitating this kind of inspection have been commented out. This is another feature I would have liked to have had the ability to toggle on and off from a terminal, but again, this feature was cut out due to time constraints.
			\par Testing correctness of the Q Table update function was achieved through gathering and plotting the metrics described in section \ref{inst}. Because it is very difficult to look at the printout of a Q Table -- all of its keys and values enumerated in sequence -- and be able to tell correctness from incorrectness, it was necessary instead to perform a more qualitative kind of verification in this case. Results from this type of testing are shown and discussed later on in section \ref{testingresults}.
			\par Automated testing for this program ended up being an insurmountable obstacle, given the scope and time constraints of the project. 
			\par I made a foray into writing properties that would be checkable with Quick Check, and was led toward Monadic checks (e.g. \texttt{forAllM}) because of the nature of my program. My goal was to write checks along the semantic lines of ``the cans and agent of any randomly-generated board should always fall inside its dimension values," and maybe ``the updated value in any Q Table should be equivalent to the Q function applied to the original value." 
			\par The barrier to the first type of check was my lack of understanding: the Quick Check examples provided in class did not address testing IO functions that I could tell, and the literature online was both wide and deep. 
			\par The barrier to the second type of check was limited time and scope, from an engineering standpoint. Writing such tests manually, without the assistance of an automated tool such as Quick Check, would have taken a lot of time that I didn't have and would have been at least as error prone as the functions they were meant to test.

			\subsubsection{Behavior \& Performance}

			\par I was interested in comparing my new program with my previously written working program, written in Python. I knew that the Python version was correct (or at least sufficiently correct to receive an A in the Machine Learning course), so a comparison between the data gathered from both programs would help me get an idea of whether my new program was correct as well. Additionally, I was curious to see how the two programs compared in terms of time required to execute the same training/testing sequence.
			\par Both programs report their results by generating a graphical plot in the form of a pdf or png. This would be the way in which I would gather information about the behavior of each program. See these results and discussion in sections \ref{data} and \ref{interp}.
			\par For comparing performance, I conducted some tests by running each program from a bash shell composed with a call to the bash utility \texttt{time}. The programs were both run on the same 2016 Macbook with a dual core Intel m3 processor and 8 GB of RAM. The Haskell program was first compiled using GHC. Both programs train an agent on a 10x10 board (where walls make up the outermost cells) for 5,000 episodes of 200 steps, then test the trained agent for 5,000 episodes of 200 steps. Here are the initial results:\\

			\begin{lstlisting}[numbers=none]
Codys-MacBook:qlearning cody$ time ./main
Training...
Training finished.
Testing...
Testing finished.

real	0m48.034s
user	0m47.207s
sys	0m0.701s
Codys-MacBook:qlearning cody$

Codys-MacBook:qlearn cody$ time python qlearn.py

real	1m37.250s
user	1m36.285s
sys	0m0.739s
Codys-MacBook:qlearn cody$
			\end{lstlisting}\ \\

			\par It would appear that the Haskell program is much faster than the Python. However, we have a wildcard in this comparison: namely, the library calls performing the image generation. Maybe the library calls are skewing the comparison? This seemed to warrant running both programs again without the image generation step at the very end:\\

			\begin{lstlisting}[numbers=none]

Codys-MacBook:qlearning cody$ time ./main
Training...
Training finished.
Testing...
Testing finished.

real	0m43.698s
user	0m42.876s
sys	0m0.608s
Codys-MacBook:qlearning cody$

Codys-MacBook:qlearn cody$ time python qlearn.py

real	1m39.152s
user	1m38.211s
sys	0m0.715s
Codys-MacBook:qlearn cody$
					\end{lstlisting}\ \\

					\par This seems to have not made any difference. 
					\par Maybe it comes as no surprise that a compiled program runs faster than an interpreted one. And it could very well be that some of the design choices I made in the Python version were not the best. I've never had formal instruction on Python, after all. And maybe comparing the performance of the two languages in this way tells us very little, since the programs were not written using the same abstractions or algorithms. Still, the results are interesting and encouraging -- using Haskell has paid off in terms of performance, in this case!

		\subsection{Data Collected}\label{data}

			\par As discussed above, the primary metric of concern when running this program was the rewards accumulated over all training episodes, and the average rewards obtained from each testing episode. These were then plotted on a graph. Results varied a bit from experiment to experiment, but here is a typical example:\\

			\includegraphics[width=\textwidth]{rewards_eps0_1}

		\subsection{Interpretation of Data}\label{interp}

		\par At the top of the image, the average reward acquired over all testing episodes is listed. This gives us some idea of how well the agent did in a given episode during testing -- in this case we might say that it was able to pick up around 12 of the 30-ish cans on the randomized board of any given episode. This might not be totally accurate, as it does not account for ``mistakes" such as attempting pickups without cans being present, or attempting to move into walls, but it gives us some idea.
		\par The plotted line shows the sum of all values stored in the Q Table at the end of a given training episode. In some ways this gives us an idea of when most of the learning takes place during training -- we see that the algorithm learns most of the particulars of its policy by the 1,000th episode, and merely finetunes thereafter. 
			\par From this information, we can infer that this very basic Q-Learning algorithm is working correctly. However, these metrics give us only a basic idea of what is going on with the algorithm. Truly understanding the model generated or derived by a machine learning algorithm remains an area of open research \cite{nguyen}.

@@

		\par The biggest hurdle to getting this program working, from the perspective of my first attempt, was having a good understanding of IO with Haskell and Monads in general. The solution to that hurdle, of course, was to take the Functional Languages class, and learn about how Monads are themselves like programs that produce outputs, or potentially have side effects. 
		\par Additionally, it was my goal to design the program in a way appropriate to the functional paradigm. I don't think that I accomplished this during my first attempt: using heavily parametrized types as stand-ins for struct-like objects and attempting to ``loop" using explicit recursion were my two big failures here. They made my job as a functional programmer more difficult, and in some ways occluded the way toward a functionally-appropriate answer. Redesigning the program with a better understanding of Haskell helped me to keep the functional idioms in mind and ultimately to have an easier time of finding solutions to the small hangups I encountered.
		\par Finally, the ``hump," as it were, was finding a way to pass the output of one Monadic function as non-monadic input to another, accomplished using the Kleisli operator. Thanks to Professor Jones for his guidance here! After this problem was solved, writing the rest of the program was easy.

@@

	\pagebreak
	\begin{thebibliography}{9}
		\bibitem{RL}
		Melanie  Mitchell.
		\textit{Reinforcement Learning}.
		http://web.cecs.pdx.edu/~mm/MachineLearningWinter2017/RL.pdf
		\bibitem{alphago}
		Silver, D., A. Huang, Aja, C. Maddison, A. Guez, L. Sifre, G. van den Driessche, J. Schrittwieser, I. Antonoglou, V. Panneershelvam, M. Lanctot, S. Dieleman, D. Grewe, J. Nham, N. Kalchbrenner, I. Sutskever, T. Lillicrap, M. Leach, K. Kavukcuoglu, T. Graepel, D. Hassabis. 2016. "Mastering the Game of Go with Deep Neural Networks and Tree Search". Nature 529 (28 Jan): 484--503.
		\bibitem{kleisli}
		Monad laws.
		https://wiki.haskell.org/Monad\_laws
		\bibitem{nguyen}
		Nguyen A, Yosinski J, Clune J. 
		\textit{Deep Neural Networks are Easily Fooled: High Confidence Predictions
for Unrecognizable Images.}
		In Computer Vision and Pattern Recognition (CVPR ?15), IEEE, 2015.
	\end{thebibliography}
